---
title: "Basic data manipulation language (DML)"
author: "Kevin Wang"
date: "20 Nov 2022"
output: html_document
---

# Data preparation in `R`

Evaluated only once to create a CSV file.

```{r, eval = FALSE}
library(palmerpenguins)
library(dplyr)
library(readr)
library(lubridate)
readr::write_csv(x = penguins, file = "data/penguins.csv")
```

# Setting up connection (in `RSQlite`)

This chunk of code will take the in-memory data and allow this data to be queried exactly like a SQL database. 

The print methods for `R` is different to that of `SQL`. 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
penguins = readr::read_csv(file = "data/penguins.csv")
dbListTables(con)

dbWriteTable(con, "penguins", penguins)
dbListTables(con)
dbListFields(con, "penguins")
```

# Selecting specific columns

```{sql, connection=con}
SELECT species, island 
FROM penguins
```

## Select all columns

```{sql, connection=con}
SELECT * 
FROM penguins;
```

## Selecting distinct entries from 

```{sql, connection=con}
SELECT DISTINCT species 
FROM penguins;
```

```{sql, connection=con}
SELECT DISTINCT species, island 
FROM penguins;
```

# Counting number of distinct entries

```{sql, connection=con}
SELECT species, 
       COUNT(species) AS n
FROM penguins
GROUP BY species;
```

# Sorting

```{sql, connection=con}
SELECT species, 
       COUNT(*)*1.0/(SELECT COUNT(*) AS total FROM penguins) AS proportion
FROM penguins
GROUP BY species 
ORDER BY proportion DESC
```


# Limited rows

## Filtering for the first few rows

```{sql, connection=con}
SELECT *
FROM penguins
LIMIT 10
```

## Filtering for the first few rows ordered by some variable

```{sql, connection=con}
SELECT * 
FROM penguins
ORDER BY body_mass_g DESC
LIMIT 10
```


# Filtering

## Filtering for rows

`WHERE` is the keyword for handling both numeric and character columns. It should evaluate to some boolean value. 

## Numeric columns

```{sql, connection=con}
SELECT species, 
       COUNT(species) AS n
FROM penguins
WHERE body_mass_g > 4000
GROUP BY species
```

The `BETWEEN` keyword is useful.

```{sql, connection=con}
SELECT species, 
       COUNT(species) AS n
FROM penguins
WHERE body_mass_g 
BETWEEN 4000 AND 4500
GROUP BY species
```

## Categorical columns

The major difference is that R/Python compared to SQL is that the former use double equal sign `==` whereas `SQL` uses a single equal sign `=` when subsetting categorical variables. 

```{sql, connection=con}
SELECT species, 
       COUNT(species) AS n
FROM penguins
WHERE island = 'Biscoe'
GROUP BY species
```

```{sql, connection=con}
SELECT species, 
       COUNT(species) AS n
FROM penguins
WHERE island IN ("Biscoe", "Dream")
GROUP BY species
```

# Missing values 

`R` uses `NA` to encode missing values. 

`SQL` uses `NULL` to encode missing values, which is not to be confused with `NULL` in `R` or `Python`.

```{sql, connection=con}
SELECT species,
       COUNT(species) AS n
FROM penguins
WHERE sex IS NULL
GROUP BY species
```

```{sql, connection=con}
SELECT species,
       COUNT(species) AS n
FROM penguins
WHERE sex IS NOT NULL
GROUP BY species
```

# String operations


SQL's LIKE operator is a powerful tool for pattern matching in string comparison. It's especially useful for filtering rows based on specific text patterns within columns. Here are some examples:

+ The percent symbol (`%`) represents any sequence of zero or more characters. 
+ The underscore (`_`) is used to match exactly one character. 
+ Using square brackets `[]`, you can match any single character within the specified set.


The first task aims to tally the species of penguins, for entries with island begin with the characters "Torg".

```{sql, connection=con}
SELECT species, island, COUNT(species) AS n
FROM penguins
WHERE island LIKE "Torg%"
GROUP BY species
```

The second task aims to tally the species of penguins, for entries with island's second character equal to "r".

```{sql, connection=con}
SELECT species, island, COUNT(species) AS n
FROM penguins
WHERE island LIKE "_r%"
GROUP BY species
```

# Summarisation

```{sql, connection=con}
SELECT species, 
       MIN(body_mass_g) AS min_mass,
       AVG(body_mass_g) AS mean_mass,
       MAX(body_mass_g) AS max_mass,
       STDEV(body_mass_g) AS sd_mass,
       AVG(body_mass_g/1000.0) AS mass_kg
FROM penguins
GROUP BY species
```


# Proportion calculation

In the code below, we are calculating the frequencies of penguins and the percentages. Notice that this percentage is calculated with respect to the entire penguin population, i.e. the total is needed. 

In `SQL`, the output will be of same variable type as the input, so the multiplication of `1.0` is necessary to make the outputs come out as decimals rather than being rounded to an integer.

```{sql, connection=con}
SELECT species, 
       COUNT(*)*1.0/SUM(COUNT(*)) OVER () AS proportion
FROM penguins
GROUP BY species 
ORDER BY proportion
```

Alternatively:

```{sql, connection=con}
SELECT species, 
       COUNT(*)*1.0/(SELECT COUNT(*) AS total FROM penguins) AS proportion
FROM penguins
GROUP BY species 
ORDER BY proportion DESC
```

# Filtering using summary statistics

In `SQL`, the `WHERE` function cannot operate on summary statistics, this is a job for `HAVING`.

In the task below, we will group by the species and sex of the penguins and filter for penguins species/sex combo that has an average mass of greater than 4700 grams. The result should be Gentoo male penguins. 

```{sql, connection=con}
SELECT *
FROM penguins
WHERE body_mass_g IS NOT NULL
GROUP BY species, sex
HAVING AVG(body_mass_g) >= 4700;
```
