---
title: "Joining datasets"
author: "Kevin Wang"
date: "20 Nov 2022"
output: html_document
---

# Data preparation 

Evaluated only once to create a CSV file.

```{r, eval = FALSE}
library(nycflights13)
library(dplyr)
library(readr)
library(lubridate)
set.seed(123)
readr::write_csv(x = nycflights13::airports %>% 
                   dplyr::select(faa, name), 
                 file = "data/airports.csv")
readr::write_csv(x = nycflights13::airlines, file = "data/airlines.csv")
readr::write_csv(x = nycflights13::flights %>% 
                   dplyr::slice(1:100) %>% 
                   dplyr::select(carrier, flight, tailnum, origin, dest), 
                 file = "data/flights.csv")
```

This chunk of code will take the in-memory data and allow this data to be queried exactly like a SQL database. 

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
airports = readr::read_csv(file = "data/airports.csv")
airlines = readr::read_csv(file = "data/airlines.csv")
flights = readr::read_csv(file = "data/flights.csv")
dbListTables(con)

dbWriteTable(con, "airports", airports)
dbWriteTable(con, "airlines", airlines)
dbWriteTable(con, "flights", flights)
dbListTables(con)
dbListFields(con, "airports")
dbListFields(con, "airlines")
dbListFields(con, "flights")
```

There are three data sets in this page: airports, airlines and flights. 

# Inner join

`SQL` needs explicitly specify the common column by name on both data.

```{sql, connection=con}
SELECT *
FROM flights
INNER JOIN airlines
  ON flights.carrier = airlines.carrier
```

# AS: Aliasing table names

We can freely rename the naming of variables in `R`. 

```{sql, connection=con}
SELECT *
FROM flights AS f
INNER JOIN airlines AS a
  ON f.carrier = a.carrier
```

# Multiple inner joins 

One of the annoying thing is that `SQL` allows columns with the same name, so some manual renaming is necessary to avoid confusion in the final output. We will do this in the next section.

In the following example, flight data is extracted. Each flight is then joined with the `airlines` table via the `carrier` column, and also joined with the `airports` table with the `origin` = `faa` specification. 

```{sql, connection=con}
SELECT *
FROM flights
INNER JOIN airlines
  ON flights.carrier = airlines.carrier
INNER JOIN airports
  ON flights.origin = airports.faa
```

# Renaming/subquery during join 

Sometimes, joining entire tables can be inefficient, especially if not all columns are needed. Thus, during the inner join call, `SELECT ... FROM` can be invoked within brackets and this subquery can also be aliased. 

```{sql, connection=con}
SELECT *
FROM flights
INNER JOIN (SELECT carrier, name AS carrier_name FROM airlines) AS rap
  ON flights.carrier = rap.carrier
INNER JOIN (SELECT faa, name AS airport_name FROM airports) AS ral
  ON flights.origin = ral.faa
```

# Case when

Case when allows for alternate coding for some categorical information. 

```{sql, connection=con}
SELECT *, 
  CASE WHEN origin = "JFK" THEN 4
       WHEN origin = "EWR" THEN 3
       WHEN origin = "LGA" THEN 2
       ELSE 0 END
       AS runways
FROM flights;
```

# SELECT INTO to create a temporary table 

Unfortunately, the in-memory `RSQLite` set-up for this worksheet does not allow me to execute this SQL command. 

```{sql, eval=FALSE}
-- Creating frequency of carrier summary table
SELECT carrier,
       COUNT(carrier) AS n
INTO freq_carrier
FROM flights
GROUP BY carrier
ORDER BY n DESC;
-- Joining that with the airline data
INNER JOIN airlines ON freq_carrier.carrier = airlines.carriers
```
