---
title: "Subqueries"
author: "Kevin Wang"
date: "20 Nov 2022"
output: html_document
---

# Data preparation 

Evaluated only once to create a CSV file.

```{r, eval = FALSE}
library(nycflights13)
library(dplyr)
library(readr)
library(lubridate)
set.seed(123)
readr::write_csv(x = nycflights13::airports %>% 
                   dplyr::select(faa, name), 
                 file = "data/airports.csv")
readr::write_csv(x = nycflights13::airlines, file = "data/airlines.csv")
readr::write_csv(x = nycflights13::flights %>% 
                   dplyr::slice(1:100) %>% 
                   dplyr::mutate(date = paste(year, month, day, sep = "/") %>% lubridate::ymd()) %>% 
                   dplyr::select(date, carrier, flight, arr_delay, dep_delay, origin, dest), 
                 file = "data/flights_dates.csv")
```

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(DBI)
# Create an ephemeral in-memory RSQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
airports = readr::read_csv(file = "data/airports.csv")
airlines = readr::read_csv(file = "data/airlines.csv")
flights = readr::read_csv(file = "data/flights_dates.csv")
dbListTables(con)

dbWriteTable(con, "airports", airports)
dbWriteTable(con, "airlines", airlines)
dbWriteTable(con, "flights", flights)
dbListTables(con)
dbListFields(con, "airports")
dbListFields(con, "airlines")
dbListFields(con, "flights")
```

# Subqueries

SQL queries can be nested. The subquery can be any part of `SELECT`, `FROM`, `WHERE` and `GROUP BY`.

The main ways that you can use a subquery: 

1. Comparing groups to summarised values
2. Reshaping data
3. Combine data that cannot be joined

## Subquery in `WHERE`

In the following example, we first check for the average departure delays in the `flights` data, across all carriers. 

```{sql, connection=con}
SELECT AVG(dep_delay) FROM flights
```
We can then use this average value to filter for flights that are greater than the average departure delays. 

```{sql, connection=con}
SELECT carrier, dep_delay
FROM flights
WHERE dep_delay > (SELECT AVG(dep_delay) FROM flights)
```

We can further summarise the average delays by carrier, restricting to only the 33 records above. 

```{sql, connection=con}
SELECT carrier, AVG(dep_delay) AS carrier_avg_dep_delay
FROM flights
WHERE dep_delay > (SELECT AVG(dep_delay) FROM flights)
GROUP BY carrier
```

Note that, the `WHERE` clause has a higher order of operation than `SELECT.`. Thus, the result we generate is the carrier-specific results of the 33 records, and NOT the result restricted to only carriers that are greater than the average. To illustrate this, pay attention to the query below. We see that AA has an average departure delay of -0.58, which is less than the overall average of -0.23. But in the query above shows a result of 5.6 for AA. This is because the WHERE clause is executed first before SELECT.

```{sql, connection=con}
SELECT carrier, AVG(dep_delay) AS carrier_avg_dep_delay
FROM flights
GROUP BY carrier
```

Here is the general order of operations in an SQL query:

1. FROM clause: This clause specifies the tables that the query will operate on.
1. WHERE clause: This clause specifies the conditions that must be met for a row to be included in the query results.
1. GROUP BY clause: This clause groups the rows of the query result into sets based on the values of one or more columns.
1. HAVING clause: This clause filters the grouped rows based on a specified condition.
1. SELECT clause: This clause specifies the columns that will be included in the query results.
1. ORDER BY clause: This clause specifies the order in which the rows of the query result will be sorted.

## Subquery in the `FROM` statement

<!-- This is a case where we can transform the data from long to wide format before selecting. -->

Note that this code will return an error, because that the `WHERE` clause can not operate on a variable that is being created in the `SELECT` clause. Again, because that the `SELECT` clause has a lower order of operation than `WHERE`, this is why `WHERE` cannot operate on a variable that has not been created yet. 

```{sql, connection=con, error=TRUE}
SELECT carrier, AVG(dep_delay) AS avg_dep_delay
FROM flights as f
WHERE avg_dep_delay > 0
GROUP BY carrier
```

This problem can be solved by the `HAVING` clause as below. 

```{sql, connection=con, error=TRUE}
SELECT carrier, AVG(dep_delay) AS avg_dep_delay
FROM flights as f
GROUP BY carrier
HAVING avg_dep_delay > 0
```

However, we can also turn the query as a subquery and then perform additional operations on it. 

```{sql, connection=con}
SELECT carrier, avg_dep_delay
FROM (
  SELECT carrier, AVG(dep_delay) AS avg_dep_delay
  FROM flights as f
  GROUP BY carrier)
WHERE avg_dep_delay > 0
```

## Subquery inside the `SELECT` clause

Here, we need to return a single value to compare against individual values. 

```{sql, connection=con}
SELECT carrier, COUNT(carrier) AS num_flights
FROM flights
GROUP BY carrier
```


# Improve readability using the `WITH` statement

<!-- ```{sql, connection=con} -->
<!-- SELECT f.carrier, AVG(f.dep_delay) AS mean_dep_delay -->
<!-- FROM flights AS f -->
<!-- INNER JOIN airlines AS a -->
<!--   ON f.carrier = a.carrier -->
<!-- GROUP BY f.carrier -->
<!-- ``` -->

In the example below, a full SQL query summarising the departure delay as a table, `t`. Another operation is performed on this summarised table to extract the maximum average departure delay carrier.

```{sql, connection=con}
WITH t as (SELECT f.carrier, a.name, AVG(f.dep_delay) AS mean_dep_delay
      FROM flights AS f
      INNER JOIN airlines AS a
        ON f.carrier = a.carrier
      GROUP BY f.carrier)
SELECT *
FROM t
WHERE t.mean_dep_delay = (SELECT MAX(t.mean_dep_delay) FROM t);
```
## `WITH` can be used on multiple tables 

The only thing needed here is to use a comma to separate the different tables. 

```{sql, connection=con}
WITH 
  nyc_airports AS 
    (SELECT faa, name
     FROM airports
     WHERE faa IN ("LGA", "JFK", "EWR")),
  other_airports AS 
    (SELECT * FROM airports)
SELECT carrier, origin, dest, nyc_airports.name AS origin_name, other_airports.name AS dest_name
FROM flights
INNER JOIN nyc_airports
ON flights.origin = nyc_airports.faa
INNER JOIN other_airports
ON flights.dest = other_airports.faa
```

# `CASE WHEN`

This function is rather straight-forward. The syntax generally follows that: CASE WHEN logical THEN encoding ELSE encoding END.

In general, it changes how variables are encoded and will allow us to perform aggregation.

```{sql, connection=con}
SELECT dep_delay, arr_delay
FROM flights as f
```


```{sql, connection=con}
SELECT CASE WHEN dep_delay <= 0 AND arr_delay >= 0 THEN 'left early, arrived late'
            WHEN dep_delay >= 0 AND arr_delay <= 0 THEN 'left late, arrived early'
            WHEN dep_delay <= 0 AND arr_delay <= 0 THEN 'left early, arrived early'
            WHEN dep_delay >= 0 AND arr_delay >= 0 THEN 'left late, arrived late'
            ELSE NULL END AS delay_patterns
FROM flights as f
```


```{sql, connection=con}
SELECT 
  SUM(CASE WHEN dep_delay <= 0 AND arr_delay >= 0 THEN 1 ELSE 0 END) AS left_early_arrived_late,
  SUM(CASE WHEN dep_delay >= 0 AND arr_delay <= 0 THEN 1 ELSE 0 END) AS left_late_arrived_early,
  SUM(CASE WHEN dep_delay <= 0 AND arr_delay <= 0 THEN 1 ELSE 0 END) AS left_early_arrived_early,
  SUM(CASE WHEN dep_delay >= 0 AND arr_delay >= 0 THEN 1 ELSE 0 END) AS left_late_arrived_late
FROM flights as f
```


# Correlated subqueries 

Correlated subqueries are subqueries that reference one or more columns in the main query. Correlated subqueries depend on information in the main query to run, and thus, cannot be executed on their own.

Correlated subqueries are evaluated in SQL once per row of data retrieved, a process that takes a lot more computing power and time than a simple subquery.

In the example below, the subquery in the WHERE clause will first calculate the average arrival delay for each carrier. This information is computed carrier by carrier and only the flights with with the arrival delay less than that carrier's average are returned for the main query.  

```{sql, connection=con}
SELECT 
  main.carrier, 
  main.arr_delay
FROM flights as main
WHERE 
  main.arr_delay < 
  (
    SELECT AVG(sub.arr_delay)
    FROM flights as sub
    WHERE main.carrier = sub.carrier
  )
```

This information is, in theory, identical to this query, which is more spread out and explicit in how the JOIN operation is used. 

Using a join instead of a subquery can be more efficient because the database engine can optimize the join operation and potentially use indexes on the carrier column. In contrast, a subquery may require the database engine to perform a separate query for each row in the main table, which can be slow for large datasets.

```{sql, connection=con}
SELECT 
  main.carrier, 
  main.arr_delay
FROM flights as main
JOIN 
  (
    SELECT carrier, AVG(arr_delay) as avg_delay
    FROM flights
    GROUP BY carrier
  ) as sub
ON 
  main.carrier = sub.carrier
  AND main.arr_delay < sub.avg_delay
```
